---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

options(width = 300)
```

# HL7.NCIMS.R

<!-- badges: start -->
<!-- badges: end -->

The goal of HL7.NCIMS.R is to enable wrangling of HL7 2.3.1/2.4 in R. This package was
driven by the need of HL7 wrangling for NCIMS tasks (a notifiable conditions database for 
a state health authority). HL7 messages are just flattened list objects, and as such 
can be imported into R as a nested list.

Information on HL7 specifications were sourced from:  

- http://www.hl7.eu/HL7v2x/v231/hl7v231segm.htm
- https://hl7-definition.caristix.com/v2/

## Installation

### Devops

Clone repo, open `HL7.NCIMS.R.Rproj` and use `devtools`

```{r}
library(devtools)
load_all()
```

### From archive

Someone built the package and has given you the file - be aware if they built using 
source or binary. Generally, `tar.gz` = source, `zip` = binary

```{r, eval = F}
install.packages('HL7.NCIMS.R.xxx.tar.gz', repos = NULL, type = 'source')
install.packages('HL7.NCIMS.R.xxx.zip', repos = NULL)
```

### From DRAT

A local package repo may exist. Again, be aware if they built using source or binary.
Look for `./bin` and `./src` in the DRAT folder.

```{r, eval = F}
library(drat)
drat::addRepo("workgroup", 'file:drive:/path/to/drat')
install.packages('HL7.NCIMS.R', repos = options()$repos[2]) # assuming binary
```

## Parsing a HL7

The first problem this package solves is parsing of HL7 messages into a nested list. The 
function `parse_hl7_message()` will parse a HL7 file into a list. This works for both
single messages and batch messages. For each parsed message, the function will try it's
best at naming all the elements. This is simple for the top level elements (HL7 segments), 
but a bit more cumbersome within the list elements (HL7 segment fields). By convention, fields are 
referenced by number ("The fourth field in MSH is the sending facility"), but
having names can make things a bit more readable at times. Repeated segments within a
message are numbered in the output list e.g. OBX.1 OBX.2 (using the Set ID value).  

**Notes**

- All values should be imported as text, and any conversion needs to be done downstream e.g with datetimes
- A `filename` attribute is attached with the list object

### Single HL7 message  

Use `parse_hl7_message()` on a file path.

```{r}
hl7_file <- system.file(package = 'HL7.NCIMS.R', 'extdata/hl7-2.3.1.hl7')

hl7_list <- parse_hl7_message(hl7_file)

names(hl7_list)

# First 5 fields of these segments
hl7_list$MSH[1:5]
hl7_list$PID[1:5]
```

```{r}
# File from where data parsed
attr(hl7_list, 'filename')
```

Traditionally indexes are used, but using named elements can help readability.

```{r}
with(hl7_list,
     data.frame(
       first_name = PID[[5]][[2]],
       last_name = PID[[5]][[1]],
       suburb = PID[[11]][[3]]
     )
)

with(hl7_list,
     data.frame(
       first_name = PID$PatientName$givenName,
       last_name = PID$PatientName$familyName,
       suburb = PID$PatientAddress$city
     )
)
```

### Batch HL7 message

`parse_hl7_message()` will check for batch headers and parse appropriately. The
result is a list of parsed messages.

```{r}
hl7_file <- system.file(package = 'HL7.NCIMS.R', 'extdata/fake-covid-batch.hl7')

hl7_list <- parse_hl7_message(hl7_file)

# unnamed list of length of the number of messages
length(hl7_list)
names(hl7_list)

# a single message lives within the list elements now
names(hl7_list[[1]])

# accessing things
hl7_list[[1]]$PID$PatientName
hl7_list[[2]]$PID$PatientName
```

### Multiple HL7 to line list

Cycle through each file and use `parse_hl7_message()`

```{r}
# Two HL7 files starting with 'fake' are distributed with this package, 
hl7_files <- 
  system.file(package = 'HL7.NCIMS.R', 'extdata') %>% 
  list.files(pattern = 'fake-covid-\\d.hl7$', full.names = T)

basename(hl7_files)

# Parse into list
hl7_list <- lapply(hl7_files, parse_hl7_message)

length(hl7_list)

# Equivalent result to a parsed batch HL7
hl7_list[[1]]$PID$PatientName
hl7_list[[2]]$PID$PatientName

# el = list element in each loop
lapply(hl7_list, function(el){
  
  with(el,
       data.frame(
         first_name = PID$PatientName$givenName,
         last_name = PID$PatientName$familyName,
         suburb = PID$PatientAddress$city,
         lab = MSH$SendingFacility,
         test = OBX.1$ObservationIdentifier$text,
         result = OBX.1$ObservationValue[[2]]
       )
  )
}) %>% 
  do.call(rbind.data.frame, .)
```

## Creating HL7 2.3.1

The second problem this package solves is converting any arbitrary piece of data into a HL7 message.
This is conducted by piecing together _*segments*_ such as the message header (MSH), Patient identification (PID)
and Observation/Results (OBX). This package provides segment functions e.g. `MSH()`, `PID()`, with named parameters that will 
build a text segment ready for piecing together using the function `build_hl7()`.

**Notes**  

- There a default blank values for all the fields in a segment function, and they will be included 
in the final output as shown below up until the last observed value i.e. trailing blanks are trimmed
see the parameter `.trim` in most functions
- There are helper functions `*Components()` for nested fields. As these are helpers, you can safely skip
if you can correctly create the required value
- Dates/Datetimes can be converted using `datetime_to_hl7_datetime()`


```{r}
example_hl7_build <-
  build_hl7(
    MSH(SendingFacility = 'A lab', ReceivingFacility = 'NSW Health', VersionID = '2.3.1'),
    PID(PatientID = '1234', 
        PatientName = PatientNameComponents(familyName = 'Ross', 
                                            givenName = 'Bob'), 
        PatientAddress = PatientAddressComponents(streetAddress = '123 Fake Street', 
                                                  city = 'Springfield',
                                                  zipPostcode = '90120')),
    OBX(SetID = 1,
        ValueType = 'CE',
        ObservationIdentifier = ObservationIdentifierComponents(identifier = 'NSW_LOINC-376',
                                                                text = 'nCoV-2019 PCR',
                                                                nameOfCodingSystem = 'LN'),
        ObservationValue = variedComponents('260415000^Not Detected^SNOMED-CT'))
  )
```

```{r}
example_hl7_build
```

Conversion of Date and Datetimes

```{r}
datetime_to_hl7_datetime(Sys.time())
```

`.trim` will trim trailing blank fields. It is `TRUE` by default

```{r}
MSH(SendingFacility = 'NSW HEALTH', VersionID = '2.3.1')
MSH(SendingFacility = 'NSW HEALTH', VersionID = '2.3.1', .trim = FALSE)

PatientAddressComponents(streetAddress = '123 Fake Street')
PatientAddressComponents(streetAddress = '123 Fake Street', .trim = FALSE)
```

### Line list to HL7 

Often we have the task of translating a line list into HL7 messages. A simple
example is show below:

- loop through the rows using `lapply`
- convert the row into a list for easy reference e.g. `d$firstname`

```{r, echo = F, eval = F}
some_line_list <-
  rbind.data.frame(
    list(firstname = 'Homer', lastname = 'Simpson',
         street = '742 Evergreen Terrace', suburb = 'Springfield',
         test_id = 'ncov19', text_text = 'nCoV-2019 PCR',
         result_code = 'P',  result_text = 'Postive', 
         facility = 'Dr Hibbert Medical')
    ,
    list(firstname = 'Ned', lastname = 'Flanders',
         street = '744 Evergreen Terrace', suburb = 'Springfield',
         test_id = 'ncov19', text_text = 'nCoV-2019 PCR',
         result_code = 'N', result_text = 'Negative',
         facility = 'Nick Riviera Appartment')
  )

readr::write_csv(some_line_list, file = 'inst/extdata/fake-covid-n2.csv')
```

```{r}
# A 2 row line list with fake data is distributed with this package
some_line_list <- read.csv(system.file(package = 'HL7.NCIMS.R', 'extdata/fake-covid-n2.csv'))
some_line_list
```

```{r}
hl7_build_list <-
  lapply(1:nrow(some_line_list), function(row){
    
    # d = data element (the row, as a list)
    d <- as.list(some_line_list[row,])
    
    build_hl7(
      MSH(SendingFacility = d$facility, ReceivingFacility = 'NSW Health', VersionID = '2.3.1'),
      PID(PatientID = '1', 
          PatientName = PatientNameComponents(familyName = d$firstname, 
                                              givenName = d$lastname), 
          PatientAddress = PatientAddressComponents(streetAddress = d$street, 
                                                    city = d$suburb)),
      OBX(SetID = 1,
          ValueType = 'CE',
          ObservationIdentifier = ObservationIdentifierComponents(identifier = d$test_id,
                                                                  text = d$text_text,
                                                                  nameOfCodingSystem = 'LN'),
          ObservationValue = variedComponents(d$result_code, d$result_text))
    )
  })
```

```{r}
hl7_build_list
```

```{r, eval = F}
# for filenames 001, 002, etc if necessary
n_leading_zero <- sprintf('%%0%sd', nchar(length(hl7_build_list)))

# Output somewhere
# * note this was used to make fake-covid-*.hl7 used above
for(i in seq_along(hl7_build_list)){ 
  i_leading_zero = sprintf(fmt = n_leading_zero, i)
  filename = sprintf('Some-filename-%s.hl7', i_leading_zero)
  path = file.path('some/path', filename)
  writeLines(hl7_build_list[[i]], con = path)
}
```




