---
title: "HL7-to-linelist-Gonorrhoea"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{HL7-to-linelist-Gonorrhoea}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options(width = 300)
```

This vignette provides an example HL7 to linelist example.

```{r setup}
library(HL7.NCIMS.R)
```

## The task

Antibiotic resistance results needed extraction from HL7 messages. There were multiple 
results within each message. The goal was to create a table with 1 row per result 
along with personal and sample identifiers.

## The data

There are 2 Gonorrhoea HL7 files containing antibiotic resistance results within this package.

```{r}
gonorrhoea_files <-
  system.file('extdata', package = 'HL7.NCIMS.R') %>% 
  list.files(pattern = 'gonorrhoea', full.names = TRUE)

gonorrhoea_files
```

## Parsing

The messages were parsed using `HL7.NCIMS.R`.

```{r}
gonorrhoea_parsed_list <-
  lapply(gonorrhoea_files, parse_hl7_message)

# 2 elements (files), with 15 elements (segments) each
length(gonorrhoea_parsed_list)
sapply(gonorrhoea_parsed_list, length)
```

## Data-wrangling

This section explains several steps that were later combined into a single wrangle function.
Thus, we begin with a single parsed HL7.

```{r}
msg <- gonorrhoea_parsed_list[[1]]
```

### Extraction function

We expect the possibility of completely missing fields from parsing. For sake of 
example, let us pretend there could be a 5th element in `PatientName`. Selection
of this element when missing produces an error

```{r, error=TRUE}
msg$PID$PatientName[[5]]
```

What we want is for it to return a 'blank' value. The function `purrr::pluck`
allows for a default value when a list element is not found

```{r}
purrr::pluck(msg$PID$PatientName, 5)
purrr::pluck(msg$PID$PatientName, 5, .default = "")
```

or we can roll our own

```{r}
my_pluck <- function(x, .default = ''){
  tryCatch(x, error = function(e) '')
}

my_pluck(msg$PID$PatientName[[5]])
# and correct otherwise
my_pluck(msg$PID$PatientName[[2]])
```

regardless, to save on typing, we create an alias function.

```{r}
# p = purrr::pluck with .default = ''
p <- function(x, ...){ 
  purrr::pluck(x, ..., .default = '') 
}
```

### Personal identifiers

Personal identifiers occur only once in the message. We can safely extract them
into a `pid` data.frame object. Note the use of `with()` to further save on typing.

```{r}
pid <-
  with(msg$PID, {
    data.frame(
      first_name	= p(PatientName, 2),
      middle_name = p(PatientName, 3),
      last_name = p(PatientName, 1),
      street_address = p(PatientAddress, 1),
      suburb = p(PatientAddress, 3),
      state = p(PatientAddress, 4),
      postcode = p(PatientAddress, 5),
      dob = DateTimeOfBirth,
      gender = Sex)
  })

pid
```

### Sample number and date

Simple extraction from the observation request segment `OBR`.

```{r}
obr <-
  with(msg$OBR, {
    data.frame(specimen_number = p(FillerOrderNumber, 1),
               specimen_date = p(RequestedDatetime, 1)
    )
  })

obr
```

### Results data

The part is complex. The data is held among multiple `OBX` segments. We need to go
through them and extract certain parts, as well as conditionally output values depending
on what we find.

```{r}
obxs <- grep('OBX', names(msg))

obxs # indexes of OBX in message
```

#### Sample site

Extract the sample site from the following OBX segment:

`OBX|1|ST|M6706^Site^NATA1234|1|Urethra||||||F||0|202203171808+1000|JOHNWICK`  

We expect the segment to have a `x$ValueType == 'ST' & x$ObservationIdentifier[[2]] == 'Site'`.
If this is found extract `p(x$ObservationValue, 1)` (with `x` being the selected segment)
otherwise output `""`. Note I've used element names, but positions also work:
`x[[2]] == 'ST' & x[[3]][[2]] == 'Site'`, and `x[[5]][[1]]`.


```{r}
obxs_st_site <-
  sapply(msg[obxs], function(x){
    x$ValueType == 'ST' & x$ObservationIdentifier[[2]] == 'Site'
  })

if (any(obxs_st_site)) {
  msg_site_ind <- obxs[obxs_st_site]
  obx_site <- data.frame(site = p(msg[[msg_site_ind]]$ObservationValue, 1))
} else {
  obx_site <- data.frame(site = NA_character_)
}

obx_site
```

#### Antibiotic resistance results

Again, we look for results and act upon them if found, from

`OBX|7|CE|18906-8^Ciprofloxacin Susc Islt^LN^M20007^Ciprofloxacin^NATA1234|7|S^Susceptible^L|...`
`OBX|8|CE|18895-3^Ceftriaxone Susc Islt^LN^M20062^Ceftriaxone^NATA1234|8|S^Susceptible^L|...`

The step returns `NULL` for segments without any hits, so we remove them.

```{r}
antibiotics_OR_grep <- "Penicillin|Ciprofloxacin|Ceftriaxone|Azithromycin"

obx_tmp <-
  lapply(msg[obxs], function(x){
    
    # TODO could examine code in position 1
    # TODO could match values rather than string manipulation
    obx_id <- p(x$ObservationIdentifier, 2)
    
    if (grepl(pattern = antibiotics_OR_grep, x = obx_id, ignore.case = TRUE) ){
      
      if (grepl(' Susc ', obx_id)) {
        
        data.frame(antibiotic = sub(' .*', '', obx_id),
                   type = 'susceptibility',
                   value = p(x$ObservationValue, 1))
        
      } else if (grepl('MIC', obx_id)) {
        
        data.frame(antibiotic = sub(' .*', '', obx_id),
                   type = 'mic',
                   value = p(x$ObservationValue, 1))
      }
      
    }
  })

is.not.null <- function(x) { !is.null(x) }

obx_not_nulls <- sapply(obx_tmp, is.not.null)

obx_tmp[obx_not_nulls]
```

And further wrangle the results - row bind, complete results for 2 test types, and 
do some conversions. Some tests had categorical outputs (susceptibility categories), while others
we numerical (MIC) and needed conversion.

```{r}
categorise_etest_mic <- function(x){
  x <- sub('<', '', x)
  x <- as.numeric(x)
  dplyr::case_when(
    x < 0.002 ~ 'Susceptible',
    x < 0.05 ~ 'Less Susceptible',
    TRUE ~ 'Resistant'
  )
}

if (any(obx_not_nulls)){
  obx_result <-
    obx_tmp[obx_not_nulls] %>%
    do.call(dplyr::bind_rows, .) %>%
    tidyr::complete(type = c('susceptibility', 'mic')) %>%
    tidyr::pivot_wider(names_from = 'type', values_from = 'value') %>%
    dplyr::mutate(susceptibility = ifelse(test = is.na(susceptibility) & !is.na(mic),
                                          yes = categorise_etest_mic(mic),
                                          no = susceptibility))
} else {
  obx_result <- data.frame(antibiotic = NA, susceptibility = NA, mic = NA)
}

obx_result
```

### Completing the linelist

```{r}
msh <- data.frame(lab = p(msg$MSH$SendingFacility, 1))

cbind.data.frame(file = attr(msg, 'filename'), pid, obr, obx_site, obx_result, msh)
```

```{r, echo = F}
rm(
  pid,
  obr,
  obx_site,
  obx_result,
  msh,
  obx_tmp,
  msg,
  is.not.null,
  obxs_st_site,
  obxs,
  obx_not_nulls,
  msg_site_ind,
  antibiotics_OR_grep,
  p
)
```

### As a complete function

Being happy with the code above, they were combined into a single function.

```{r}
extract_pid_and_antibiotic_data <- function(msg, antibiotics = c('Penicillin', 'Ciprofloxacin', 'Ceftriaxone', 'Azithromycin')){
  
  antibiotics_OR_grep <- paste0(antibiotics, collapse = "|")
  
  p <- function(x,...){  purrr::pluck(x,...,.default = '') }
  
  pid <-
    with(msg$PID, {
      data.frame(
        first_name	= p(PatientName, 2),
        middle_name = p(PatientName, 3),
        last_name = p(PatientName, 1),
        street_address = p(PatientAddress, 1),
        suburb = p(PatientAddress, 3),
        state = p(PatientAddress, 4),
        postcode = p(PatientAddress, 5),
        dob = DateTimeOfBirth,
        gender = Sex)
    })
  
  obr <-
    with(msg$OBR, {
      data.frame(specimen_number = p(FillerOrderNumber, 1),
                 specimen_date = p(RequestedDatetime, 1)
      )
    })
  
  obxs <- grep('OBX', names(msg))
  
  obxs_st_site <-
    sapply(msg[obxs], function(x){
      x$ValueType == 'ST' & x$ObservationIdentifier[[2]] == 'Site'
    })
  
  if (any(obxs_st_site)) {
    msg_site_ind <- obxs[obxs_st_site]
    obx_site <- data.frame(site = p(msg[[msg_site_ind]]$ObservationValue, 1))
  } else {
    obx_site <- data.frame(site = NA_character_)
  }
  
  # malformed HL7?
  if (is.na(obx_site$site)){
    obxs_st_site <-
      sapply(msg[obxs], function(x){
        x$ObservationIdentifier[[2]] == 'Site'
      })
    if (any(obxs_st_site)) {
      msg_site_ind <- obxs[obxs_st_site]
      warning('This message did not have ValueType = ST but the site ',
              p(msg[[msg_site_ind]]$ObservationValue, 1), ' was found.')
    }
  }
  
  obx_tmp <-
    lapply(msg[obxs], function(x){
      
      # TODO could examine code in position 1
      # TODO could match values rather than string manipulation
      obx_id <- p(x$ObservationIdentifier, 2)
      
      if (grepl(pattern = antibiotics_OR_grep, x = obx_id, ignore.case = TRUE) ){
        
        if (grepl(' Susc ', obx_id)) {
          
          data.frame(antibiotic = sub(' .*', '', obx_id),
                     type = 'susceptibility',
                     value = p(x$ObservationValue, 1))
          
        } else if (grepl('MIC', obx_id)) {
          
          data.frame(antibiotic = sub(' .*', '', obx_id),
                     type = 'mic',
                     value = p(x$ObservationValue, 1))
        }
        
      }
    })
  
  
  is.not.null <- function(x) { !is.null(x) }
  
  obx_not_nulls <- sapply(obx_tmp, is.not.null)
  
  if (any(obx_not_nulls)){
    obx_result <-
      obx_tmp[obx_not_nulls] %>%
      do.call(dplyr::bind_rows, .) %>%
      tidyr::complete(type = c('susceptibility', 'mic')) %>%
      tidyr::pivot_wider(names_from = 'type', values_from = 'value') %>%
      dplyr::mutate(susceptibility = ifelse(test = is.na(susceptibility) & !is.na(mic),
                                            yes = categorise_etest_mic(mic),
                                            no = susceptibility))
  } else {
    obx_result <- data.frame(antibiotic = NA, susceptibility = NA, mic = NA)
  }
  
  msh <- data.frame(lab = p(msg$MSH$SendingFacility, 1))
  #msh = data.frame(lab = 'LAB_33', test = '698-1')
  
  cbind.data.frame(file = attr(msg, 'filename'), pid, obr, obx_site, obx_result, msh)
  
}

```

### Running the function

```{r}
gonorrhoea_parsed_list %>% 
  lapply(extract_pid_and_antibiotic_data) %>% 
  do.call(dplyr::bind_rows, .)
```
